Variance reduction for Markov chains via martingale representations

### 2020-04-14 aap/Revision/vr-mcmc-revision-new-new.tex
update

 where $\Psi_m$ is a class of polynomials
 \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}}, \quad \alpha_{\mathbf{s}}\in \mathbb{R}.
Then using the identity

@@

we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,

@@

\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}

@@

&=&

@@

\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
where for any integer $k,s$ $P_{k,s}$
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. We estimate $Q_{r}$ only for $r < n_{\mathrm{trunc}}$ where the truncation level \(n_{\mathrm{trunc}}\) may depend on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.

@@

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-\frac{\|x-\mu\|^2}{2}} + \rme^{-\frac{\|x+\mu\|^2}{2}}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the cost-variance ratio
\begin{eqnarray}
\label{eq:cvR}
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]}
\end{eqnarray}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. Since for fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to computing function $f$, we set for $K = 1$
 \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
 \end{equation*}
 since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Note that in this case each $a_{r,\mathbf{k}}(x)$  is a linear function, which can be computed at the same cost as $f$. Similarly, for $K = 2$ we set
  \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8
 \end{equation*}
since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.

@@

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
\end{eqnarray}
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
\par

>>>>

 where $\Psi_m$ is a class of polynomials $ \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}}$ with $\alpha_{\mathbf{s}}\in \mathbb{R}$. Then using the identity

@@

we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x=(x^1,\dots,x^d) \in \mathbb{R}^d$,

@@

\widehat a_{r,\mathbf{k}} (x) &= \int \mathbf{H}_\mathbf{k}(z) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}z) P_\xi(\rmd z)

@@

&=

@@

\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x^i - \gamma \mu_i(x)),
where for any integer $k,s$ $P_{k,s}$ is a one-dimensional polynomial of degree at most \(s\) with analytically known coefficients. We estimate $Q_{r}$ only for $r < n_{\mathrm{trunc}}$ where the truncation level \(n_{\mathrm{trunc}}\) may depend on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.

@@

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-(1/2)\|x-\mu\|^2} + \rme^{-(1/2)\|x+\mu\|^2}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the cost-variance ratio
\begin{eqnarray}
\label{eq:cvR}
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]}
\end{eqnarray}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. Since for fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to computing function $f$, we set for $K = 1$
 \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
 \end{equation*}
 since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Note that in this case each $a_{r,\mathbf{k}}(x)$  is a linear function, which can be computed at the same cost as $f$. Similarly, for $K = 2$ we set
  \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8
 \end{equation*}
since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.

@@

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-(1/2)(x-\mu)^T\Sigma^{-1}(x-\mu)} + \rme^{-(1/2)(x+\mu)^T\Sigma^{-1}(x+\mu)}\right)
\end{eqnarray}
where $\rho \in (0,1)$, $\mu \in \rset^2$, $\Sigma$ is a symmetric positive definite $2 \times 2$ matrix and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
\par
